Deriving Random Generators via\\ Higher-Level Representations

### 2019-01-20 paper/main.tex
WIP

% Local Variables:
% TeX-master: "main.lhs.tex"
% TeX-command-default: "Make"
% End:

@@

\newcommand{\evalrep}[2]{\ensuremath{\llbracket #1 \rrbracket_{#2}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Some parameters
\setlength{\mathindent}{1.5\parindent}

@@

The main contribuitions of this paper are:
%
\begin{itemize}
  %
\item We identify two patological scenarios for which standard type-driven
  automatic derivation tools fail to synthesize practical random generators, due
  to a lack of either type structure or domain knowleadge (Section
  \ref{sec:randomtesting}).
  %
\item We present a generation technique able to encode stronger properties of
  the target data type by reifying the static information present on the program
  codebase (Section \ref{sec:hrep}).
  %
\item We apply and extend the theory of branching processes to analitically
  predict the average distribution of generated values.
  %
  Furthermore, we use the predictions to perform simulation-based optimization
  of the random generation parameters (Section \ref{sec:synthesis}).
  %
\item We provide an implementation of our ideas in the form of a Haskell library
  to perform automatic derivation of random generators capable to extract useful
  structure information from the user source code.
  %
\end{itemize}
\section{Type-Driven Random Generation} \label{sec:randomtesting}

In this section we briefly describe the common technique used for automatically
deriving random data generators in Haskell using a type-driven approach.
%
Then, we introduce its main drawbacks by considering two scenarios where this
technique gives poor results in practice.
%
We remark that, altough this work makes focus on Haskell data types, this
technique is general enough to be applied to most programming languages with
some level of support for composite types.

Haskell is a strongly typed programming language with a powerful type system.
%
It lets programmers encode a considerable amount of information about the
properties of their systems using data types that can be checked at compilation
time.
%
One of its key aspects is the support for Algebraic Data Types (ADTs).
%
Essentially, an ADTs is a composite type defined by combining other types.
%
In the most basic conception, types can be combined by \textbf{sums} (also known
as \emph{variant types} or \emph{tagged disjoint unions}) and \textbf{products}
(or tuples) of other data types.
%
To exemplify this, consider the following definition of the data type \ensuremath{\Conid{Exp}}
encoding integer expressions:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;\Conid{Exp}{}\<[11]%
\>[11]{}\mathrel{=}{}\<[15]%
\>[15]{}\Conid{Val}\;{}\<[20]%
\>[20]{}\Conid{Int}{}\<[E]%
\\
\>[11]{}\;\;\vert\;{}\<[15]%
\>[15]{}\Conid{Add}\;{}\<[20]%
\>[20]{}\Conid{Exp}\;{}\<[25]%
\>[25]{}\Conid{Exp}{}\<[E]%
\\
\>[11]{}\;\;\vert\;{}\<[15]%
\>[15]{}\Conid{Mul}\;{}\<[20]%
\>[20]{}\Conid{Exp}\;{}\<[25]%
\>[25]{}\Conid{Exp}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

In the previous definition, we declare \ensuremath{\Conid{Exp}} as the sum of three possible
classes of values: \ensuremath{\Conid{Val}} represents literal values, whereas \ensuremath{\Conid{Add}} and \ensuremath{\Conid{Mul}}
represent the addition and multiplication of two integer expressions,
respectively.
%
In Haskell, \ensuremath{\Conid{Val}}, \ensuremath{\Conid{Add}} and \ensuremath{\Conid{Mul}} are called \emph{data constructors} (or
constructors for short) and are used to distiguish which variant of the data
type we are reffering to.
%
Each data constructor is then defined as a product of zero or more types known
as \emph{fields}.
%
In particular, \ensuremath{\Conid{Val}} contains a field of type \ensuremath{\Conid{Int}}, while \ensuremath{\Conid{Add}} and \ensuremath{\Conid{Mul}}
contain two fields of type \ensuremath{\Conid{Exp}} representing the operands of each operation.
%
Note that \ensuremath{\Conid{Exp}} is used as a field of at least one of its data constructors
(case \ensuremath{\Conid{Add}} and \ensuremath{\Conid{Mul}}), making it a recursively defined ADT.
%
In general, we will say that a data constructor with no recursive fields is
\emph{terminal}, and \emph{non-terminal} or \emph{recursive} if it has at least
one field of such nature.
%
With this reprentation, the expression ``2 + (5 * 6)'' can be encoded simply by
using \ensuremath{\Conid{Exp}} data constructors as \ensuremath{(\Conid{Add}\;(\Conid{Val}\;\mathrm{2})\;(\Conid{Mul}\;(\Conid{Val}\;\mathrm{5})\;(\Conid{Val}\;\mathrm{6})))}.
%
Furthermore, an evaluation function from \ensuremath{\Conid{Exp}}s to integer values can be defined
very idiomatically:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
\column{27}{@{}>{\hspre}c<{\hspost}@{}}%
\column{27E}{@{}l@{}}%
\column{30}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{eval}\mathbin{::}\Conid{Exp}\to \Conid{Int}{}\<[E]%
\\
\>[B]{}\Varid{eval}\;(\Conid{Val}\;\Varid{n}){}\<[17]%
\>[17]{}\mathrel{=}\Varid{n}{}\<[E]%
\\
\>[B]{}\Varid{eval}\;(\Conid{Add}\;\Varid{x}\;\Varid{y}){}\<[17]%
\>[17]{}\mathrel{=}\Varid{eval}\;\Varid{x}{}\<[27]%
\>[27]{}\mathbin{+}{}\<[27E]%
\>[30]{}\Varid{eval}\;\Varid{y}{}\<[E]%
\\
\>[B]{}\Varid{eval}\;(\Conid{Mul}\;\Varid{x}\;\Varid{y}){}\<[17]%
\>[17]{}\mathrel{=}\Varid{eval}\;\Varid{x}{}\<[27]%
\>[27]{}\mathbin{*}{}\<[27E]%
\>[30]{}\Varid{eval}\;\Varid{y}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

In the previous definition, \ensuremath{\Varid{eval}} is described using \emph{pattern matching}
over each possible kind of value.
%
For the input case of a literal values, we simply return the value contained in
the \ensuremath{\Conid{Val}} constructor.
%
On the other hand, if the input value matches either an \ensuremath{\Conid{Add}} or a \ensuremath{\Conid{Mul}} data
constructor, where \ensuremath{\Varid{x}} and \ensuremath{\Varid{y}} are \emph{pattern variables} that match any
subexpression, we recursively evaluate these subexpressions, combining them with
the apropiate operation on each case.

@@

In order to perform random testing of a Haskell codebase involving user defined
data types, most approaches require the user to provide a random data generator
for each one of them.
%
This tends to be a cumbersome and error prone task that closely follows the data
type structure.
%
For instance, consider the following definition of a \quickcheck random
generator for the type \ensuremath{\Conid{Exp}}:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
\column{6}{@{}>{\hspre}c<{\hspost}@{}}%
\column{6E}{@{}l@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}gen_{Exp}\mathrel{=}\Varid{sized}\;(\lambda \Varid{size}\to {}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\mathbf{if}\;\Varid{size}\equiv \mathrm{0}{}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\mathbf{then}\;\Conid{Val}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{arbitrary}{}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\mathbf{else}\;\Varid{frequency}{}\<[E]%
\\
\>[4]{}\hsindent{2}{}\<[6]%
\>[6]{}[\mskip1.5mu {}\<[6E]%
\>[9]{}(\mathrm{1},\Conid{Val}{}\<[18]%
\>[18]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{arbitrary}){}\<[E]%
\\
\>[4]{}\hsindent{2}{}\<[6]%
\>[6]{},{}\<[6E]%
\>[9]{}(\mathrm{2},\Conid{Add}{}\<[18]%
\>[18]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{smaller}\;gen_{Exp}\mathop{\mathsmaller{\langle \ast \rangle}}\Varid{smaller}\;gen_{Exp}){}\<[E]%
\\
\>[4]{}\hsindent{2}{}\<[6]%
\>[6]{},{}\<[6E]%
\>[9]{}(\mathrm{1},\Conid{Mul}{}\<[18]%
\>[18]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{smaller}\;gen_{Exp}\mathop{\mathsmaller{\langle \ast \rangle}}\Varid{smaller}\;gen_{Exp})\mskip1.5mu]){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 %$

This random generator is defined using \quickcheck's combinator \ensuremath{\Varid{sized}} to
paremetrize the generation process up to an external integer number known as the
\emph{generation size}.
%
This parameter is used to limit the maximum amount of recursive call that this
random generator can perform.
%
When called with a positive generation size (case \ensuremath{\Varid{gen}\;\Varid{n}}), the generator will
pick among any data constructor of \ensuremath{\Conid{Exp}} with a explicitly given frequency---two
times more \ensuremath{\Conid{Add}}s than \ensuremath{\Conid{Val}} or \ensuremath{\Conid{Mul}}.
%
In the case it picks to generate a \ensuremath{\Conid{Val}} data constructor, it also generates a
random \ensuremath{\Conid{Int}} value using the standard overloaded generator \ensuremath{\Varid{arbitrary}}
(\quickcheck provides standard random generators for most base types like \ensuremath{\Conid{Int}},
\ensuremath{\Conid{Bool}}, etc.).
%
On the other hand, when it randomly picks to generate either an \ensuremath{\Conid{Add}} or a \ensuremath{\Conid{Mul}}
data constructor, it also generates independently two random subexpressions
corresponding to the data constructor fields by calling itself recursively
(\ensuremath{\Varid{resize}\;(n{-}1)\;gen_{Exp}}), decreasing the generation size on each recursive
invocation.

This procedure keeps calling itself recursively until the generation size
reaches zero (case \ensuremath{\Varid{gen}\;\mathrm{0}}), where the generator is constrained to pick only
among terminal data constructors, being \ensuremath{\Conid{Val}} the only possible choice in our
particular case.
%
Strictly decreasing the the generation size by one on each recursive call
results in a useful property over the generated data: every generated value has
at most \ensuremath{\Varid{n}} levels, where \ensuremath{\Varid{n}} is the generation size set by the user.
%
This property enables us to model the generation process using the theory of
branching processes introduced by \tocite, and extended in this work as
described in Section \ref{sec:synthesis}.

Having discussed the previous random generator definition, it easy to understand
how to extend this generation procedure to any data type defined in an algebraic
fashion.
%
Fortunately, there exists different meta-programming tools to avoid the user
from having to mechanically write random generators over and over again for each
user-defined ADT.
%
The simplest tool for such purpose is \megadeth.
%
Given the name of the target data type, it synthesizes a random generator for it
that behaves pretty much like the previously defined one.
%
However, picking among different data constructors with uniform probability can
lead to a generation process biased to generate (in average) very small values,
regardless of the generation sized set by the user \tocite.

\dragen is meta-programming tool conceived to mitigate this problem.
%
Instead of setting a uniform generation probability of data constructors, this
tool considers the scenario where each one can be generated following a
different (although fixed over time) generation probability.
%
Then, this tool uses the theory of branching processes to analitically predict
the average distribution of data constructors generated on each random value.
%
This prediction mechanism is used to feedback a simulation-based optimization
process that adjusts the generation probability of each data constructor in
order to obtain a particular distribution of values that can be specified by the
user, offering a more flexible testing environment while still being mostly
automated.

Both \megadeth and \dragen synthesize random generators that are theoretically
capable to generate the whole space of values of the target data type, provided
that we set a generation size suficiently big.
%
However, the limitations arise quickly when we consider that the underlying
generation model is essentially the same: they pick a single data constructor
and recusively generate each required subexpression independently.
%
In practice, this procedure is too generic to generate complex data that is
useful enough to be used for random testing.

In this work we identify two sources of additional structure information which
are not considered by the aforementioned automatic derivation tools to obtain
random data generators:

\begin{enumerate}
\item The target code behaves differently over inputs matching very specific
  patterns.
\item The target code yields part of the responsability of preserving its
  invariants to the functions of its abstract interface.
\end{enumerate}

This information can be reified in compile time and used to synthesize richer
random generators automatically. We proceed to exemplify the previous points in
detail.

@@

To exemplify the firts problematic scenarion that may arise, suppose we want to
use randomly generated \ensuremath{\Conid{Exp}}s to test a property comprising the following
function:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{foo}\mathbin{::}\Conid{Exp}\to \Conid{Exp}{}\<[E]%
\\
\>[B]{}\Varid{foo}\;(\Conid{Add}\;(\Conid{Add}\;\Varid{x}\;(\Conid{Val}\;\mathrm{50}))\;(\Conid{Add}\;(\Conid{Val}\;\mathrm{25})\;\Varid{y})){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\mathrel{=}\Varid{error}\;\text{\ttfamily \char34 pattern~\#1\char34}{}\<[E]%
\\
\>[B]{}\Varid{foo}\;(\Conid{Mul}\;(\Conid{Val}\;\mathrm{50})\;(\Conid{Mul}\;(\Conid{Val}\;\Varid{x})\;\Varid{y})){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\mathrel{=}\Varid{error}\;\text{\ttfamily \char34 pattern~\#2\char34}{}\<[E]%
\\
\>[B]{}\Varid{foo}\;\Varid{x}\mathrel{=}\Varid{x}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

This function behaves very much like an identity function, with the exception
that it fails for two very specific patterns of input values defined using
nested pattern matching.
%
This is, \ensuremath{\Varid{foo}} does not only matches against the root data constructor, but also
against the data constructors of its subexpressions and sub-subexpressions.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% HRep

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Constructor names

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% gen

% Local Variables:
% TeX-mbigstarer: "main.lhs.tex"
% TeX-command-default: "Make"
% End:

@@

\newcommand{\hrep}[1]{\ensuremath{H\!Rep_{#1}}}
\newcommand{\gen}[1]{\ensuremath{gen_{#1}}}
\newcommand{\term}[1]{\ensuremath{#1^\bigstar}}
\newcommand{\evalrep}[2]{ \ensuremath{\llbracket #1 \rrbracket_{#2}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Some parameters
\setlength{\mathindent}{\parindent}

@@

We remark that, altough this work focuses on Haskell data types, this technique
is general enough to be applied to most programming languages with some level of
support for composite types.

The main contribuitions of this paper are:
%
\begin{itemize}
  %
\item We identify two patological scenarios for which standard type-driven
  automatic derivation tools fail to synthesize practical random generators, due
  to a lack of either type structure or domain knowleadge (Section
  \ref{sec:randomtesting}).
  %
\item We present a generation technique able to encode stronger properties of
  the target data type by reifying the static information present on the program
  codebase (Section \ref{sec:hrep}).
  %
\item We apply and extend the theory of branching processes to analitically
  predict the average distribution of generated values.
  %
  Furthermore, we use the predictions to perform simulation-based optimization
  of the random generation parameters (Section \ref{sec:synthesis}).
  %
\item We provide an implementation of our ideas in the form of a Haskell library
  to perform automatic derivation of random generators capable to extract useful
  structure information from the user source code.
  %
\end{itemize}
\section{Random Data Generation in Haskell} \label{sec:randomtesting}

In this section we briefly introduce the common approach for automatically
deriving random data generators in Haskell using a type-driven approach, along
with its main drawbacks.

Haskell is a strongly typed programming language with a powerful type system.
%
It lets programmers encode a considerable amount of information about the
structure of their systems using data types that can be checked at compilation
time.
%
One of its key aspects is the support for Algebraic Data Types (ADTs).
%
Essentially, an ADTs is a composite type defined by combining other types in
terms of \textbf{sums} (also known as \emph{variant types}) and
\textbf{products} (or tuples) of other data types.
%
To exemplify this, consider the following type definition to encoding \ensuremath{\Conid{Html}}
expressions:
%
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{12}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
\column{30}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;\Conid{Html}{}\<[12]%
\>[12]{}\mathrel{=}{}\<[16]%
\>[16]{}\Conid{Text}\;{}\<[22]%
\>[22]{}\Conid{String}{}\<[E]%
\\
\>[12]{}\,\ \vert\;{}\<[16]%
\>[16]{}\Conid{Sing}\;{}\<[22]%
\>[22]{}\Conid{String}{}\<[E]%
\\
\>[12]{}\,\ \vert\;{}\<[16]%
\>[16]{}\Conid{Join}\;{}\<[22]%
\>[22]{}\Conid{Html}\;{}\<[30]%
\>[30]{}\Conid{Html}{}\<[E]%
\\
\>[12]{}\,\ \vert\;{}\<[16]%
\>[16]{}\Conid{Tag}\;{}\<[22]%
\>[22]{}\Conid{String}\;{}\<[30]%
\>[30]{}\Conid{Html}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

In the previous definition, we declare \ensuremath{\Conid{Html}} as the sum of four possible
constructions: \ensuremath{\Conid{Text}} represents plain text values. \ensuremath{\Conid{Sing}} and \ensuremath{\Conid{Tag}} represent
singular and paired Html tags, and \ensuremath{\Conid{Join}} concats two expression one after
another.
%
We only encode a very small subset of the actual Html specification for
illustrative reasons.
%
In Haskell, \ensuremath{\Conid{Text}}, \ensuremath{\Conid{Sing}} and \ensuremath{\Conid{Join}} and \ensuremath{\Conid{Tag}} are known as \emph{data
  constructors} (or constructors for short) and are used to distiguish which
variant of the data type we are constructing.
%
Each data constructor is defined as a product of zero or more types known as
\emph{fields}.
%
For instance, \ensuremath{\Conid{Text}} has a field of type \ensuremath{\Conid{String}}, whereas \ensuremath{\Conid{Join}} two recusive
fields of type \ensuremath{\Conid{Html}}.
%
In general, we will say that a data constructor with no recursive fields is
\emph{terminal}, and \emph{non-terminal} or \emph{recursive} if it has at least
one field of such nature.
%
With this reprentation, the expression ``\ensuremath{<\!\!html\!\!>\!\!hello\!\!<\!\!hr\!\!>\!\!bye\!\!<\!\!/html\!\!>}'' can be encoded as:
%
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Conid{Tag}\;\text{\ttfamily \char34 html\char34}\;(\Conid{Join}\;(\Conid{Join}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}(\Conid{Text}\;\text{\ttfamily \char34 hello\char34})\;(\Conid{Sing}\;\text{\ttfamily \char34 hr\char34}))\;(\Conid{Text}\;\text{\ttfamily \char34 bye\char34})){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
%
Additionally, we can define a function \ensuremath{\Varid{render}} to serialize \ensuremath{\Conid{Html}} values as
follows:
%
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{render}\mathbin{::}\Conid{Html}\to \Conid{String}{}\<[E]%
\\
\>[B]{}\Varid{render}\;(\Conid{Text}\;\Varid{t})\mathrel{=}\Varid{t}{}\<[E]%
\\
\>[B]{}\Varid{render}\;(\Conid{Sing}\;\Varid{t})\mathrel{=}\text{\ttfamily \char34 <\char34}\plus \Varid{t}\plus \text{\ttfamily \char34 >\char34}{}\<[E]%
\\
\>[B]{}\Varid{render}\;(\Conid{Join}\;\Varid{x}\;\Varid{y})\mathrel{=}\Varid{render}\;\Varid{x}\plus \Varid{render}\;\Varid{y}{}\<[E]%
\\
\>[B]{}\Varid{render}\;(\Conid{Tag}\;\Varid{t}\;\Varid{x}){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\mathrel{=}\text{\ttfamily \char34 <\char34}\plus \Varid{t}\plus \text{\ttfamily \char34 >\char34}\plus \Varid{render}\;\Varid{x}\plus \text{\ttfamily \char34 </\char34}\plus \Varid{t}\plus \text{\ttfamily \char34 >\char34}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

In the previous definition, \ensuremath{\Varid{render}} is described using \emph{pattern matching}
over each possible kind of value.
%
Using pattern matching we can define functions idiomatically by defining
different function clauses for each input pattern we are interested on.
%
Patterns can be defined to match specific constructors, literal values or
variable subexpressions (like \ensuremath{\Varid{t}}, \ensuremath{\Varid{x}} and \ensuremath{\Varid{y}} in the definition of \ensuremath{\Varid{render}}).
%
They can also be nested in order to match very specific patterns of values.

@@

In order to generate random value of types involving user defined ADTs, most
approaches require the user to provide a random data generator for each one of
them.
%
This is a cumbersome and error prone task that closely follows the data type
structure.
%
For instance, consider the following definition of a \quickcheck random
generator for the type \ensuremath{\Conid{Html}}:
%
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
\column{6}{@{}>{\hspre}c<{\hspost}@{}}%
\column{6E}{@{}l@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}gen_{Html}\mathrel{=}\Varid{sized}\;(\lambda \Varid{size}\to {}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\mathbf{if}\;\Varid{size}\equiv \mathrm{0}{}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\mathbf{then}\;\Varid{frequency}{}\<[E]%
\\
\>[4]{}\hsindent{2}{}\<[6]%
\>[6]{}[\mskip1.5mu {}\<[6E]%
\>[9]{}(\mathrm{2},{}\<[14]%
\>[14]{}\Conid{Text}{}\<[20]%
\>[20]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{arbitrary}){}\<[E]%
\\
\>[4]{}\hsindent{2}{}\<[6]%
\>[6]{},{}\<[6E]%
\>[9]{}(\mathrm{1},{}\<[14]%
\>[14]{}\Conid{Sing}{}\<[20]%
\>[20]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{arbitrary})\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\mathbf{else}\;\Varid{frequency}{}\<[E]%
\\
\>[4]{}\hsindent{2}{}\<[6]%
\>[6]{}[\mskip1.5mu {}\<[6E]%
\>[9]{}(\mathrm{2},{}\<[14]%
\>[14]{}\Conid{Text}{}\<[20]%
\>[20]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{arbitrary}){}\<[E]%
\\
\>[4]{}\hsindent{2}{}\<[6]%
\>[6]{},{}\<[6E]%
\>[9]{}(\mathrm{1},{}\<[14]%
\>[14]{}\Conid{Sing}{}\<[20]%
\>[20]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{arbitrary}){}\<[E]%
\\
\>[4]{}\hsindent{2}{}\<[6]%
\>[6]{},{}\<[6E]%
\>[9]{}(\mathrm{3},{}\<[14]%
\>[14]{}\Conid{Join}{}\<[20]%
\>[20]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{smaller}\;gen_{Html}\mathop{\mathsmaller{\langle \ast \rangle}}\Varid{smaller}\;gen_{Html}){}\<[E]%
\\
\>[4]{}\hsindent{2}{}\<[6]%
\>[6]{},{}\<[6E]%
\>[9]{}(\mathrm{4},{}\<[14]%
\>[14]{}\Conid{Tag}{}\<[20]%
\>[20]{}\mathop{\mathsmaller{\langle \$ \rangle}}\Varid{arbitrary}\mathop{\mathsmaller{\langle \ast \rangle}}\Varid{smaller}\;gen_{Html})\mskip1.5mu]){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 %$
%
This random generator is defined using \quickcheck's combinator \ensuremath{\Varid{sized}} to
paremetrize the generation process up to an external natural number known as the
\emph{generation size}.
%
This parameter is chosen by the user, and we use to limit the maximum amount of
recursive calls that this random generator can perform.
%
When called with a positive generation size, this generator can pick to generate
among any \ensuremath{\Conid{Html}} data constructor of with a explicitly given frequency that can
be chosen by the user (2, 1, 3 and 4 for \ensuremath{\Conid{Text}}, \ensuremath{\Conid{Sing}}, \ensuremath{\Conid{Join}} and \ensuremath{\Conid{Tag}}
constructors, respectively).
%
When it picks to generate a \ensuremath{\Conid{Val}} or a \ensuremath{\Conid{Sing}} data constructor, it also
generates a random \ensuremath{\Conid{String}} value using the standard overloaded generator
\ensuremath{\Varid{arbitrary}} (\quickcheck provides standard random generators for most base types
like \ensuremath{\Conid{String}}, \ensuremath{\Conid{Int}}, \ensuremath{\Conid{Bool}}, etc.). \footnote{%
  The operators \ensuremath{\mathop{\mathsmaller{\langle \$ \rangle}}} and \ensuremath{\mathop{\mathsmaller{\langle \ast \rangle}}} are used in Haskell to combine values obtained
  from effectful computations (like calling to a random generator) and they are
  not particularly relevant for the point being made in this work.}
%$
On the other hand, when it picks to generate either an \ensuremath{\Conid{Join}} constructor, it
also generates two independent random subexpression recursively, decreasing the
generation size by a unit on each recursive invocation (\ensuremath{\Varid{smaller}\;gen_{Html}}).
%
The case of random generation of \ensuremath{\Conid{Tag}} constructors follows analogously.

This random process keeps calling itself recursively until the generation size
reaches zero, where the generator is constrained to pick only among terminal
data constructors, being \ensuremath{\Conid{Text}} and \ensuremath{\Conid{Sing}} the only possible choices in our
particular case.

% Strictly decreasing the the generation size by one on each recursive call
% results in a useful property over the generated data: every generated value has
% at most |n| levels, where |n| is the generation size set by the user.
%
% This property enables us to model the generation process using the theory of
% branching processes introduced by \tocite, and extended in this work as
% described in Section \ref{sec:synthesis}.

The previous definition is rather mechanical, except perhaps for the chosen
generation frequencies.
%
In this light, it easy to extend this procedure to any data type defined in an
algebraic fashion.
%
Fortunately, there exists different meta-programming tools to avoid the user
from having to mechanically write random generators over and over again for each
user-defined ADT.
%
The simplest tool for such purpose is \megadeth \tocite.
%
Given a target data type, it synthesizes a random generator for it that behaves
similarly to the one presented above, where the generation frequencies are
defined to be uniform across constructors.
%
However, picking among different data constructors with uniform frequency can
lead to a generation process biased towards generating (in average) very small
values, regardless of the generation sized set by the user \tocite.

\dragen is meta-programming tool conceived to mitigate this problem.
%
Instead of setting a uniform generation probability of data constructors, this
tool this tool uses the theory of branching processes to modelize and predict
analitically the average distribution of data constructors generated on each
random value.
%
This prediction mechanism is used to feedback a simulation-based optimization
process that adjusts the generation frequency of each data constructor in order
to obtain a particular distribution of values that can be specified by the user,
providing this way a more flexible testing environment while still being mostly
automated.

Althogh both \megadeth and \dragen synthesize random generators that are
theoretically capable to generate the whole space of values of the target data
type. the limitations arise quickly when we consider that the underlying
generation model is essentially the same: they pick a single data constructor
and recusively generate each required subexpression independently.
%
In practice, this procedure is often too generic to generate random data with
enough structural complexity to be used for testing purposes.

% In this work we identify two sources of additional structure information which
% are not considered by the aforementioned automatic derivation tools to obtain
% better random data generators:

In this work we identify two patological situations which are not properly
handled by the aforementioned derivation tools:

\begin{enumerate}
\item The target code behaves differently over inputs matching specific patterns
  of nested values.
\item The target code encodes a significant portion of its structure on its
  abstract interface.
\end{enumerate}

In Section \ref{sec:hrep} we show how this structural information can be used to
synthesize richer random generators automatically.
%
We proceed to exemplify the previous points in detail.

@@

To exemplify the first problematic scenario, suppose we want to use randomly
generated \ensuremath{\Conid{Htmls}}s to test a property comprising the following function:
%
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{simpl}\mathbin{::}\Conid{Html}\to \Conid{Html}{}\<[E]%
\\
\>[B]{}\Varid{simpl}\;(\Conid{Join}\;(\Conid{Text}\;\Varid{t1})\;(\Conid{Text}\;\Varid{t2})){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\mathrel{=}\Conid{Text}\;(\Varid{t1}\plus \Varid{t2}){}\<[E]%
\\
\>[B]{}\Varid{simpl}\;(\Conid{Join}\;(\Conid{Join}\;(\Conid{Text}\;\Varid{t1})\;\Varid{x})\;\Varid{y}){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\mathrel{=}\Varid{simpl}\;(\Conid{Join}\;(\Conid{Text}\;\Varid{t1})\;(\Varid{simpl}\;(\Conid{Join}\;\Varid{x}\;\Varid{y}))){}\<[E]%
\\
\>[B]{}\Varid{simpl}\;(\Conid{Join}\;\Varid{x}\;\Varid{y})\mathrel{=}\Conid{Join}\;(\Varid{simpl}\;\Varid{x})\;(\Varid{simpl}\;\Varid{y}){}\<[E]%
\\
\>[B]{}\Varid{simpl}\;(\Conid{Tag}\;\Varid{t}\;\Varid{x})\mathrel{=}\Conid{Tag}\;\Varid{t}\;(\Varid{simpl}\;\Varid{x}){}\<[E]%
\\
\>[B]{}\Varid{simpl}\;\Varid{x}\mathrel{=}\Varid{x}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

This function simplifies sequences of \ensuremath{\Conid{Text}} constructors into a single big
\ensuremath{\Conid{Text}} constructor.
%
To do so, it has to pattern match against sequences of \ensuremath{\Conid{Text}} constructors
combined by a \ensuremath{\Conid{Join}} constructor using nested patterns (see \ensuremath{\Varid{simpl}} fist and
second clauses).
%
The remaining clauses are only meant to propagate this simplification within
nested expressions.

Ideally, we would like to test each clause of the function \ensuremath{\Varid{simpl}} approximately
the same amount of time each.
%
However, each data constructor is generated independently when using either
\megadeth or \dragen, thus the probability of generating a value satisfying a
nested pattern decreases multiplicatively with the number of constructors we
pattern against to simultaneously.
%
In our tests, we found that the first two clauses of \ensuremath{\Varid{simpl}} get exercised only
approximately $1.5\%$ of the time when using \megadeth to derive a random
generator for \ensuremath{\Conid{HTML}}.
%
On the other hand, the best result we could achieve with \dragen was only able
to exercise the first and second clauses of \ensuremath{\Varid{simpl}} approximately $3\%$ and
$6\%$ of the time, respectively.
%
With both derivation tools, the most of the time was spent testing the trivial
clauses of our function, in view of they pattern match against simpler patterns
of input values.
